With demands for natural resources increasing alongside populations, and
the effects of climate change intensifying, ecosystems worldwide are
increasingly burdened with the cumulative effects of a vast array of
environmental stressors. These observations stimulate a growing demand
for ecosystem-based approaches and regional cumulative effects
assessments. Yet environmental management still overwhelmingly operates
in silos, focusing instead on single-stressor and single-species
assessments. This is particularly concerning for exploited and
endangered species whose dynamics, like that of all species, is driven
by the network of interactions structuring ecological communities and
through which the effects of stressors can propagate indirectly. A
general lack of theoretical understanding of the effects of multiple
stressors on ecological communities, and methodological and logistical
constraints explain part of this discrepancy. In this context, the
general objective of this thesis is to evaluate the cumulative effects
of climate change and human activities on the ecological communities of
the St.~Lawrence System in eastern Canada. The general hypotheses
guiding the thesis are that species interactions influence the indirect
and non-additive spread of the effects of multiple stressors through
communities and that, as such, species interactions and stressors should
be considered together in network-scale cumulative effects assessments.

The thesis is divided in three parts. In the first part (chapter 1), I
conceptualize how stressors propagate through food webs and explore
theoretically how they affect simulated 3-species motifs and food webs
of the Canadian St.~Lawrence System. We find that overlooking species
interactions invariably underestimates the effects of stressors, and
that synergistic and antagonistic effects through species interactions
are prevalent. At the scale of food webs, we find that apex predators
are generally negatively affected and mesopredators benefit from the
effects of stressors in the St.~Lawrence System, but that species
sensitivity is dependent on food web structure. The theoretical
simulations proposed in chapter 1 validate the general hypotheses of the
thesis and provide an accessible and theory-grounded framework for the
inclusion of species interactions in cumulative effects assessments.

The second part of the thesis addresses logistical and methodological
challenges for the cumulative effects assessment of climate change and
human activities on communities of the St.~Lawrence System. In chapter
2, I address the challenge of characterizing ecological interactions in
data-deficient ecosystems. I present a new unsupervised machine learning
method to predict interactions between any given set of species, given
pairwise taxonomic proximity and a collection of known empirical
interactions between marine species. Results from chapter 2 suggest that
ecological interactions can be predicted with high accuracy, which could
promote their use for environmental management. In chapter 3, I identify
environmental issues and suggest research and management priorities to
promote cumulative effects assessment and ecosystem-based management in
the St.~Lawrence System. In chapter 4, I characterize the distribution
and intensity of environmental stressors arising from human activities
and climate change in the St.~Lawrence System. Through collaborations,
existing environmental initiatives and open data portals, data-based
indicators for 22 coastal, climate, fisheries, and marine traffic
stressors were developed. Results from chapter 4 show that stressors are
widespread and that coastal areas and the Estuary, Anticosti Gyre, and
coastal areas are particularly exposed to cumulative exposure and
cumulative hotspots.

In the third part of the thesis (chapter 5), I present a cumulative
effects assessment on 193 species of the St.~Lawrence System using a
novel network-scale approach that explicitly considers ecological
interactions and indirect effects. The approach is built on the
framework presented in chapter 1 and uses results and methods from
chapters 2, 3 and 4. I compare our approach to a conventional
species-scale assessment to expose transgressive properties arising from
species interactions and uncover cumulative effects to species that
would otherwise be overlooked. Fishes and marine mammals appear
particularly prone to indirect effects from all types of stressors; this
contrasts considerably with the limited number of stressors affecting
them directly. For certain species, considering interactions may even be
the only means of assessing the effects of stressors.

In pairing theory, environmental management and computational
capabilities, my thesis shows that ecological interactions are key to
assess the effects of multiple stressors on species and proposes an
accessible approach to integrate interactions to cumulative effects
assessments. This is particularly relevant to the management of
exploited and endangered species for which we may currently ignore
significant threats by overlooking the less obvious yet no less
significant effects arising from species interactions. Developing these
holistic capabilities is essential to operationalize ecosystem-based
management.

\begin{quote}
Keywords: cumulative effects, environmental stressors, indirect effects,
biotic interactions, non-additive effects, ecological network, food web,
St.~Lawrence System, human activities, climate change.
\end{quote}
